Quantitative Auditing of AI Fairness with differentially private Synthetic Data

### 2024-12-11 paper/paper.tex
progress

% TODO: add more to intro

@@

Let $X \in \mathcal{X}$ be a discrete random variable and $p(x) := \prob{X = x}$. Its \emph{marginal Shannon entropy} is $\entropy{X} := -\sum_{x \in \mathcal{X}} p(x) \text{log} p(x)$. It quantifies the level of uncertainty of $X$. Let $Y \in \mathcal{Y}$ be another discrete random variable. Their \emph{joint Shannon entropy} is $\entropy{X,Y} := -\sum_{x \in \mathcal{X}} \sum_{y \in \mathcal{Y}} p(x, y) \text{log} p(x, y)$.

Let $X,Y$ be discrete random variables. Their \emph{mutual information} is $\mi{X}{Y} := \entropy{X} + \entropy{Y} - \entropy{X,Y}$. It quantifies the level of dependence between $X$ and $Y$.
% By definition\cite{427884}, the joint entropy is greater than or equal to the marginal entropies $\entropy{X,Y} \geq \text{max}(\entropy{X},\entropy{Y})$. Hence, we have an upper bound of mutual information $\mi{X}{Y} \leq \text{min}(\entropy{X}, \entropy{Y})$.

@@

% TODO: reiterate intro

% TODO: goal of this framework

Our auditing framework is tripartite. It consists of three parties: the data provider, the model maker, and the third-party auditor.

% TODO: write more about these paragraphs

The data provider is responsible for supplying the raw datasets which should originate from trustworthy sources, such as government agencies like a census bureau.

The model maker develops AI models. These are AI companies or research labs specialized in training and optimizing AI models.

The third-party auditor acts as an evaluator, using our tool to audit the AI models for fairness issues by combining both the datasets and the models. These may be investigative journalists or regulatory bodies.

@@

In the COMPAS example, we may select the marginal (sex,race).

For the competition, \cite{mckenna2021winning} further manually added certain cliques based on his investigation of the competition dataset. For example, they manually added the clique (sex,city,income). In addition, they would add some edges based on some sophisticated heuristics tailored to that particular dataset.

We do not consider these additional heuristics for the generality of our approach. Meanwhile, his other apporach where the auditor does access the real dataset does not fit our framework.

While synthetic data generation methods often leverage domain knowledge to improve accuracy, we chose to employ a vanilla approach with only the maximum spanning tree method without incorporating additional heuristics. This decision was made due to two main factors: first, the need for more specific domain knowledge about the dataset makes it difficult to identify meaningful dependencies; second, prediction targets may differ from dataset to dataset, which means that tailored heuristics could not be effectively generalized. By avoiding dataset-specific adjustments, our method remains general and applicable to a wide range of datasets. Furthermore, the bare tree structure makes the method more tractable and computationally efficient.

@@

In the COMPAS example, the selected the marginals are perturbed by Gaussian noise, as shown in Table~\ref{tab:marginal-perturbed}.

@@

In the COMPAS example, if the sensitive attribute is sex and the protected group is female, the protected group predicate would be $S := r(sex) = Female$. This can be easily implemented in Python as a comparison function.

@@

@@

% TODO:
first list formulas

count vars

theyre not part of marginals

they cant be computed

we can say less about fairness with these with less info

auditor need to know theyre aproxing scores

auditor need to keep in mind that these scores are for datasets and we preserve these scores thats it. scores may not be accurate. it doesnt mean its less easy to test fairness

@@

% TODO:
it works!

@@

independence
\[
P[\hat{Y} | S] - P[\hat{Y} | \overline{S}]
= \frac{P[\n{\hat{Y}} \cap \y{S}]}{P[\y{S}]} - \frac{P[\n{\hat{Y}} \cap \y{\overline{S}}]}{P[\y{\overline{S}}]}
\]

separation
\[
P[\hat{Y} | S, Y] - P[\hat{Y} | \overline{S}, Y]
= \frac{P[\n{\hat{Y}} \cap \y{S} \cap \y{Y}]}{P[\y{S} \cap \y{Y}]} - \frac{P[\n{\hat{Y}} \cap \y{\overline{S}} \cap \y{Y}]}{P[\y{\overline{S}} \cap \y{Y}]}
\]

sufficiency
\[
P[Y | S, \hat{Y}] - P[Y | \overline{S}, \hat{Y}]
= \frac{P[\y{Y} \cap \y{S} \cap \n{\hat{Y}}]}{P[\y{S} \cap \n{\hat{Y}}]} - \frac{P[\y{Y} \cap \y{\overline{S}} \cap \n{\hat{Y}}]}{P[\y{\overline{S}} \cap \n{\hat{Y}}]}
\]

accuracy equality
\[
P[Y = \hat{Y} | {S} ] - P[Y = \hat{Y} | \overline{S}]
= \frac{P[\y{Y} = \n{\hat{Y}} \cap \y{S}]}{P[\y{S}]} - \frac{P[\y{Y} = \n{\hat{Y}} \cap \y{\overline{S}}]}{P[\y{\overline{S}}]}
\]

>>>>

@@

Let $X \in \mathcal{X},Y \in \mathcal{Y}$ be discrete random variables. Their \emph{mutual information} $\mi{X}{Y}$ is
\[
\mi{X}{Y} = \sum_{x \in \mathcal{X}} \sum_{y \in \mathcal{Y}} P_{(X,Y)}(x,y) \text{log} \frac{P_{(X,Y)}(x,y)}{P_X(x)P_Y(y)}
\]
where $P_{(X,Y)}$ is the joint probability distribution of $X,Y$ and $P_X,P_Y$ are the marginal probability distributions of $X,Y$ respectively.
It quantifies the level of dependency between $X$ and $Y$.
% By definition\cite{427884}, the joint entropy is greater than or equal to the marginal entropies $\entropy{X,Y} \geq \text{max}(\entropy{X},\entropy{Y})$. Hence, we have an upper bound of mutual information $\mi{X}{Y} \leq \text{min}(\entropy{X}, \entropy{Y})$.

@@

AI systems are widely used in various aspects in our society. However, these systems may be biased against certain demographic groups. For example, COMPAS, the AI system used in American criminal justice system to predict recidivism, has been found to be biased by ProPublica, a third party media.

To ensure the fairness of AIs, third-party audits such as this is crucial. Auditing can reveal instances of unfairness in AI systems, allowing for corrective actions to be taken. We hence advocate for a framework that centers around third-party audits.

Our auditing framework is tripartite. It consists of three parties: the data provider, the model maker, and the third-party auditor. The COMPAS investigation could fit in with this framework.

The data provider is responsible for supplying the raw datasets which should originate from trustworthy sources, such as government agencies like a census bureau. In the COMPAS example, the data provider is the government agencies that provide criminal justice data, such as the Broward County Sheriff's Office in Florida.

The model maker develops AI models. These are AI companies or research labs specialized in training and optimizing AI models. The model maker in the COMPAS example is Northpointe, Inc., the company that developed the COMPAS system.

The third-party auditor acts as an evaluator, using our tool to audit the AI models for fairness issues by combining both the datasets and the models. These may be investigative journalists or regulatory bodies. ProPublica is the third-party auditor in the COMPAS example, using quantitative methods to evaluate the fairness of the COMPAS system using Florida's data.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{compas}
  \caption{Roles of each party in the COMPAS example.}
  \label{fig:compas}
  \Description{Roles of each party in the COMPAS example.}
\end{figure}

% TODO: add compas example in the following

@@

For the competition, \cite{mckenna2021winning} further manually added certain cliques based on his investigation of the competition dataset. For example, they manually added the clique (sex,city,income). In addition, they would add some edges based on some sophisticated heuristics tailored to that particular dataset. We do not consider these additional heuristics for the generality of our approach. Meanwhile, his other apporach where the auditor does access the real dataset does not fit our framework.

While synthetic data generation methods often leverage domain knowledge to improve accuracy, we chose to employ a vanilla approach with only the maximum spanning tree method without incorporating additional heuristics. This decision was made due to two main factors: first, the need for more specific domain knowledge about the dataset makes it difficult to identify meaningful dependencies; second, prediction targets may differ from dataset to dataset, which means that tailored heuristics could not be effectively generalized. By avoiding dataset-specific adjustments, our method remains general and applicable to a wide range of datasets, in addition to being more computatioanlly efficient.

@@

In the COMPAS example, we may select the marginal (sex,race). This marginal is then perturbed by Gaussian noise, as shown in Table~\ref{tab:marginal-perturbed}.

@@

@@

In the COMPAS example, the sensitive attribute is sex and the protected group is female, the protected group predicate would be $S := r(sex) = Female$; prediction is if a person will re-offend $\hat{Y} := \mathcal{M}_{recid}(r)$ where $M$ is the model's prediction of row $r$; ground truth is whether a person does re-offend $Y := r(recid)$. These can be easily implemented in Python as a comparison function.

@@

% % TODO:
% first list formulas

% count vars

% theyre not part of marginals

% they cant be computed

% we can say less about fairness with these with less info

% auditor need to know theyre aproxing scores

% auditor need to keep in mind that these scores are for datasets and we preserve these scores thats it. scores may not be accurate. it doesnt mean its less easy to test fairness

@@

% % TODO:
% it works!

@@

% independence
% \[
% P[\hat{Y} | S] - P[\hat{Y} | \overline{S}]
% = \frac{P[\n{\hat{Y}} \cap \y{S}]}{P[\y{S}]} - \frac{P[\n{\hat{Y}} \cap \y{\overline{S}}]}{P[\y{\overline{S}}]}
% \]

% separation
% \[
% P[\hat{Y} | S, Y] - P[\hat{Y} | \overline{S}, Y]
% = \frac{P[\n{\hat{Y}} \cap \y{S} \cap \y{Y}]}{P[\y{S} \cap \y{Y}]} - \frac{P[\n{\hat{Y}} \cap \y{\overline{S}} \cap \y{Y}]}{P[\y{\overline{S}} \cap \y{Y}]}
% \]

% sufficiency
% \[
% P[Y | S, \hat{Y}] - P[Y | \overline{S}, \hat{Y}]
% = \frac{P[\y{Y} \cap \y{S} \cap \n{\hat{Y}}]}{P[\y{S} \cap \n{\hat{Y}}]} - \frac{P[\y{Y} \cap \y{\overline{S}} \cap \n{\hat{Y}}]}{P[\y{\overline{S}} \cap \n{\hat{Y}}]}
% \]

% accuracy equality
% \[
% P[Y = \hat{Y} | {S} ] - P[Y = \hat{Y} | \overline{S}]
% = \frac{P[\y{Y} = \n{\hat{Y}} \cap \y{S}]}{P[\y{S}]} - \frac{P[\y{Y} = \n{\hat{Y}} \cap \y{\overline{S}}]}{P[\y{\overline{S}}]}
% \]
